taxize: taxonomic search and retrieval in R

### 2013-09-30 f1000_rev/taxize_ms.Rnw
added nfew version of ms to f1000_rev folder, folded in changes from reviewers, issue #178

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Welcome to writeLaTeX --- just edit your LaTeX on the left,
% and we'll compile it for you on the right. If you give 
% someone the link to this page, they can edit at the same
% time. See the help menu above for more info. Enjoy!
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% For more detailed article preparation guidelines, please see:
% http://f1000research.com/author-guidelines

@@

@@

\begin{abstract}

All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming increasingly available on the web, but scientists need a way to access it in a programmatic fashion that's easy and reproducible. We have developed taxize, an open-source software package (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}) for the R language. taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package facilitates open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.

\end{abstract}
\clearpage

\section*{Introduction}

Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study objects into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more.

The use of taxonomic names is, unfortunately, not straightforward. Taxonomic names often vary due to name revisions at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. For example, a study found that a compilation of 308,000 plant observations from 51 digitized herbarium records had 22,100 unique taxon names, of which only ~13,000 were accepted names \cite{weiser2007,boyle2013}. In addition, there is no one authoritative taxonomic names source for all taxa - although, there are taxon specific sources that are used by many scientists. Different sources (e.g., uBio, Tropicos, ITIS) may use different accepted names for the same taxon. For example, while the Integrated Taxonomic Information Service (ITIS) has \emph{Helianthus x glaucus} as an accepted name, The Plant List (\url{http://www.theplantlist.org}) gives that name as unresolved. But \emph{Helianthus glaucus} is an accepted name in The Plant List, while ITIS does not list this name.

One attempt to help inconsistencies in taxonomy is the use of numeric codes. For example, ITIS assigns a Taxonomic Serial Number (TSN) to each taxon, while the Universal Biological Indexer and Organizer (uBio) assigns each taxon a NameBank identifier (namebankID), and Tropicos assigns their own identifier to each taxon. Codes are helpful within a database as they can easily refer to, for example, \emph{Helianthus annuus} with a code like 123456 instead of its whole name. However, each database uses their own code; in this case for \emph{Helianthus annuus}, ITIS uses 36616, uBio uses 2658020, and Tropicos uses 40022652. As there are no universal codes for taxa across databases, this can lead to additional confusion. Last, name comparisons across databases have to be done with the actual names, not the codes.

Taxonomic data is getting easier to obtain through the web (e.g., \url{http://eol.org/}). However, there are a number of good reasons to obtain taxonomic information programatically rather than through a web interface. First, if you have more than a few names to look up on a website, it can take quite a long time to enter each name, get data, and repeat for each species. Programatically getting taxonomic names solves the problem by looping over a list of names. In addition, doing taxonomic searching, etc. becomes reproducible. With increasing reports of irreproducibility in science \cite{stodden2010,zimmer2012}, it is extremely important to make science workflows repeatable. Science workflows can now easily incorporate text, code, and images in a single executable document \cite{yihui2013}. Reproducible documents should become mainstream in biology to avoid mistakes, and make collaboration easier.

@@

The goal of taxize, an R package in development, is to make many use cases that involve retrieving and resolving taxonomic names easy and reproducible. In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would enter when interacting with a website. Therefore, we hope that moving from a web to an R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

@@

\begin{table*}[!ht]
\centering
\caption{
\bf{Some key functions in taxize, what they do, and their data sources}}
\begin{tabular}{|l|L|L|}
\hline
Function name & What it does & Source \\
\hline
apg\_lookup & Changes names to match the APGIII list & Angiosperm Phylogeny Group \url{http://www.mobot.org/MOBOT/research/APweb/}  \\
classification & Upstream classification & Various  \\
col\_children & Direct children & Catalogue of Life \url{http://www.catalogueoflife.org/}  \\
col\_downstream & Downstream taxa to specified rank & Catalogue of Life \url{http://www.catalogueoflife.org/}  \\
eol\_hierarchy & Upstream classification & Encyclopedia of Life \url{http://eol.org/}  \\
eol\_search & Search EOL taxon information & Encyclopedia of Life \url{http://eol.org/}  \\
get\_seqs & Get NCBI sequences & National Center for Biotechnology Information \url{federhen}  \\
get\_tsn & Get ITIS TSN & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
get\_uid & Get NCBI UID & National Center for Biotechnology Information \cite{federhen}  \\
searchbycommonname & Search ITIS by common name & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
searchbyscientificname & Search ITIS by scientific name & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
gisd\_isinvasive & Invasiveness status & Global Invasive Species Database \url{http://www.issg.org/database/welcome/}  \\
gni\_parse & Parse scientific names into components & Global Names Index \url{http://gni.globalnames.org/}   \\
gni\_search & Search EOL's global names index & Global Names Index \url{http://gni.globalnames.org/}   \\
gnr\_resolve & Resolve names using EOL's global names index & Global Names Resolver \url{http://resolver.globalnames.org/}  \\
itis\_downstream & Downstream taxa to specified rank & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
iucn\_status & IUCN status & IUCN Red List \url{http://www.iucnredlist.org}  \\
phylomatic\_tree & Get a plant Phylogeny & Phylomatic \cite{webb2005}  \\
plantminer & Search Plantminer & Plantminer \cite{carvalho2010plantminer}   \\
tax\_name & Get taxonomic name for specific rank & Various  \\
tax\_rank & Get rank of a taxonomic name & Various  \\
tnrs & Resolve names using iPlant & iPlant Taxonomic Name Resolution Service \url{http://tnrs.iplantcollaborative.org/}  \\
tp\_acceptednames & Check for accepted names using Tropicos & Tropicos \url{http://www.tropicos.org/}  \\
tpl\_search & Search the Plant List & The Plant List \url{http://www.theplantlist.org}  \\
ubio\_namebank & Search uBio & uBio \url{http://www.ubio.org/index.php?pagename=sample_tools}  \\
\hline
\end{tabular}
\label{tab:a}
\end{table*}

@@

\section*{Data sources and package details}

taxize uses many data sources (Table~\ref{tab:a}), and more can easily be added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomatic) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize does not venture deeply into these other domains.

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key so that they can better manage their servers, collect analytics, and shut down users that abuse the API. The services that require an API key in taxize are: Encyclopedia of Life (EOL) \url{http://eol.org/}, the Universal Biological Indexer and Organizer (uBio) \url{http://www.ubio.org/index.php?pagename=sample_tools}, Tropicos \url{http://www.tropicos.org/}, and Plantminer \cite{carvalho2010plantminer}. One can easily obtain API keys by visiting the website of each service (see Table~\ref{tab:a} for links to each site). There are two typical ways of using API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio\_namebank(srchName='Ursus americanus', key='your\_alphanumeric\_key')}). Second, you can store your key in the .Rprofile file, which is a common place to store settings. We recommend the second option as it simplifies function calls as taxize detects the stored keys.

One available data source in taxize is The Plant List \url{http://www.theplantlist.org}. The connection in taxize is done via the \emph{taxonstand} package \cite{Cayuela_2012} that solely interacts with the Plant List. We provide a few convenience functions that wrap \emph{taxonstand} into taxize.

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{install.packages}(\hlstring{"taxize"})
\hlfunctioncall{library}(taxize)
\end{alltt}
\end{kframe}
\end{scriptsize}
\end{knitrout}

Advanced users can also download and install the latest development copy from GitHub \url{https://github.com/ropensci/taxize_}.

@@

This is a common task in biology. We often have a list of species names and we want to know a) if we have the most up to date names, b) if our names are spelled correctly, and c) the scientific name for a common name. One way to resolve names is via the Global Names Resolver (GNR) service provided by the Encyclopedia of Life \url{http://resolver.globalnames.org/}. Here, we are searching for two misspelled names:

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
temp <- \hlfunctioncall{gnr_resolve}(names = \hlfunctioncall{c}(\hlstring{"Helianthos annus"}, \newline \hlstring{"Homo saapiens"}))
temp[, -\hlfunctioncall{c}(1, 4)]
\end{alltt}
\begin{verbatim}
                 matched_name       data_source_title
1        Helianthus annuus L.       Catalogue of Life
2            Helianthus annus GBIF Taxonomic Backbone
3            Helianthus annus                     EOL
4         Helianthus annus L.                     EOL
5            Helianthus annus           uBio NameBank
6 Homo sapiens Linnaeus, 1758       Catalogue of Life
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

The correct spellings are \emph{Helianthus annuus} and \emph{Homo sapiens}. Another approach uses the Taxonomic Name Resolution Service via the Taxosaurus API \url{http://taxosaurus.org/} developed by iPLant and the Phylotastic organization. In this example, we provide a list of species names, some of which are misspelled, and we'll call the API with the \emph{tnrs} function.

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
mynames <- \hlfunctioncall{c}(\hlstring{"Helianthus annuus"}, \hlstring{"Pinus contort"}, \hlstring{"Poa anua"}, \newline \hlstring{"Abis magnifica"}, \hlstring{"Rosa california"}, \hlstring{"Festuca arundinace"}, \newline \hlstring{"Sorbus occidentalos"}, \hlstring{"Madia sateva"})
\hlfunctioncall{tnrs}(query = mynames)[, -\hlfunctioncall{c}(5:7)]
\end{alltt}
\begin{verbatim}
         submittedName        acceptedName    sourceId score
9    Helianthus annuus   Helianthus annuus iPlant_TNRS     1
10   Helianthus annuus   Helianthus annuus        NCBI     1
4        Pinus contort      Pinus contorta iPlant_TNRS  0.98
5             Poa anua           Poa annua iPlant_TNRS  0.96
3       Abis magnifica     Abies magnifica iPlant_TNRS  0.96
7      Rosa california    Rosa californica iPlant_TNRS  0.99
8      Rosa california          California        NCBI     1
2   Festuca arundinace Festuca arundinacea iPlant_TNRS  0.99
1  Sorbus occidentalos Sorbus occidentalis iPlant_TNRS  0.99
6         Madia sateva        Madia sativa iPlant_TNRS  0.97
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

Another common use case is when there are many synonyms for a species. In this example, we have three synonyms of the currently accepted name for a species.

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
mynames <- \hlfunctioncall{c}(\hlstring{"Helianthus annuus ssp. jaegeri"}, \newline \hlstring{"Helianthus annuus ssp. lenticularis"}, \newline \hlstring{"Helianthus annuus ssp. texanus"}, \newline \hlstring{"Helianthus annuus var. lenticularis"}, \newline \hlstring{"Helianthus annuus var. macrocarpus"}, \newline \hlstring{"Helianthus annuus var. texanus"})
tsn <- \hlfunctioncall{get_tsn}(mynames)
\hlfunctioncall{ldply}(tsn, itis_acceptname)
\end{alltt}
\begin{verbatim}
  submittedTsn      acceptedName acceptedTsn
1       525928 Helianthus annuus       36616
2       525929 Helianthus annuus       36616
3       525930 Helianthus annuus       36616
4       536095 Helianthus annuus       36616
5       536096 Helianthus annuus       36616
6       536097 Helianthus annuus       36616
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: Integrated Taxonomic Information System (ITIS) \url{http://www.itis.gov/} and National Center for Biotechnology Information (NCBI) \cite{federhen}. First, we'll search for two species, \emph{Abies procera} and \emph{Pinus contorta} within ITIS.

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})
\hlfunctioncall{classification}(specieslist, db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}
$`Abies procera`
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Abies  18031
11       Species   Abies procera 181835

$`Pinus contorta`
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Pinus  18035
11       Species  Pinus contorta 183327
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, \newline db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}
      family
1 Asteraceae
\end{verbatim}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, \newline db = \hlstring{"ncbi"})
\end{alltt}
\begin{verbatim}
      family
1 Asteraceae
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_tsn}(searchterm = \hlstring{"Heliastes"}, searchtype = \hlstring{"sciname"})
\end{alltt}
\begin{verbatim}
           combinedname    tsn
1     Heliastes bicolor 615238
2   Heliastes chrysurus 615250
3     Heliastes cinctus 615573
4  Heliastes dimidiatus 615257
5  Heliastes hypsilepis 615273
6 Heliastes immaculatus 615639
7 Heliastes opercularis 615300
8      Heliastes ovalis 615301
 1 
NA 
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
splist <- \hlfunctioncall{c}(\hlstring{"annona cherimola"}, \hlstring{"annona muricata"}, \newline \hlstring{"quercus robur"}, \hlstring{"shorea robusta"}, \hlstring{"pandanus patina"}, \newline \hlstring{"oryza sativa"}, \hlstring{"durio zibethinus"})
\hlfunctioncall{get_tsn}(searchterm = splist, searchtype = \hlstring{"sciname"})
\end{alltt}
\begin{verbatim}
[1] "506198" "18098"  "19405"  "506787" "507376" "41976"  "506099"
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

@@

\begin{knitrout}
\begin{scriptsize}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_uid}(sciname = \hlfunctioncall{c}(\hlstring{"Chironomus riparius"}, \hlstring{"aaa vva"}))
\end{alltt}
\begin{verbatim}
[1] "315576" NA      
attr(,"class")
[1] "uid"
\end{verbatim}
\end{kframe}
\end{scriptsize}
\end{knitrout}

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, and functional trait ecology \cite{poff2006functional}. Yet, many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in Fig. \ref{fig:phylomatic}.

>>>>

@@

<<setup, include=FALSE, cache=FALSE, size='scriptsize'>>=
## set global chunk options
opts_chunk$set(fig.align='center', fig.show='hold', par=TRUE, warning=FALSE, message=FALSE, error=FALSE, comment=NA)
@

@@

\begin{abstract}
All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming increasingly available on the web, but scientists need a way to access it in a programmatic fashion that's simple and reproducible. We have developed taxize, an open-source software package for the R language (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}). taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package facilitates open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.
\end{abstract}
\clearpage

\section*{Introduction}
Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study objects into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more.

The use of taxonomic names is, unfortunately, not straightforward. Taxonomic names often vary due to name revisions at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. For example, a study found that a compilation of 308,000 plant observations from 51 digitized herbarium records had 22,100 unique taxon names, of which only ~13,000 were accepted names \cite{weiser2007,boyle2013}. In addition, there is no one authoritative source of taxonomic names for all taxa - although, there are taxon specific sources that are used by many scientists. Different sources (e.g., uBio [Universal Biological Indexer and Organizer], Tropicos, ITIS [Integrated Taxonomic Information Service]) may use different accepted names for the same taxon. For example, while ITIS has \emph{Helianthus x glaucus} as an accepted name, The Plant List (\url{http://www.theplantlist.org}) gives that name as unresolved. But \emph{Helianthus glaucus} is an accepted name in The Plant List, while ITIS does not list this name.

One attempt to help inconsistencies in taxonomy is the use of numeric codes. For example, ITIS assigns a Taxonomic Serial Number (TSN) to each taxon, while uBio assigns each taxon a NameBank identifier (namebankID), and Tropicos assigns their own identifier to each taxon. Codes are helpful within a database as they can easily refer to, for example, \emph{Helianthus annuus} with a code like 123456 instead of its whole name. However, each database uses their own code; in this case for \emph{Helianthus annuus}, ITIS uses 36616, uBio uses 2658020, and Tropicos uses 40022652. As there are no universal codes for taxa across databases, this can lead to additional confusion. Last, name comparisons across databases have to be done with the actual names, not the codes.

Taxonomic data is getting easier to obtain through the web (e.g., \url{http://eol.org/}). However, there are a number of good reasons to obtain taxonomic information programatically rather than through a web interface. First, if you have more than a few names to look up on a website, it can take quite a long time to enter each name, get data, and repeat for each species. Programatically getting taxonomic names solves the problem by looping over a list of names. In addition, doing taxonomic searching, etc. becomes reproducible. With increasing reports of irreproducibility in science \cite{stodden2010,zimmer2012}, it is extremely important to make science workflows repeatable.

@@

The goal of taxize is to make many use cases that involve retrieving and resolving taxonomic names easy and reproducible. In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would enter when interacting with a website. Therefore, we hope that moving from a web to an R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

@@

\begin{table*}[!ht]
\centering
\caption{
\bf{Some key functions in taxize, what they do, and their data sources}}
\begin{tabular}{|l|L|L|}
\hline
Function name & What it does & Source \\
\hline
apg\_lookup & Changes names to match the APGIII list & Angiosperm Phylogeny Group \url{http://www.mobot.org/MOBOT/research/APweb/}  \\
classification & Upstream classification & Various  \\
col\_children & Direct children & Catalogue of Life \url{http://www.catalogueoflife.org/}  \\
col\_downstream & Downstream taxa to specified rank & Catalogue of Life \url{http://www.catalogueoflife.org/}  \\
eol\_hierarchy & Upstream classification & Encyclopedia of Life \url{http://eol.org/}  \\
eol\_search & Search EOL taxon information & Encyclopedia of Life \url{http://eol.org/}  \\
get\_seqs & Get NCBI sequences & National Center for Biotechnology Information \cite{federhen2012}  \\
get\_tsn & Get ITIS TSN & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
get\_uid & Get NCBI UID & National Center for Biotechnology Information \cite{federhen2012}  \\
gisd\_isinvasive & Invasiveness status & Global Invasive Species Database \url{http://www.issg.org/database/welcome/}  \\
gni\_parse & Parse scientific names into components & Global Names Index \url{http://gni.globalnames.org/}   \\
gni\_search & Search EOL's global names index & Global Names Index \url{http://gni.globalnames.org/}   \\
gnr\_resolve & Resolve names using EOL's global names index & Global Names Resolver \url{http://resolver.globalnames.org/}  \\
itis\_downstream & Downstream taxa to specified rank & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
iucn\_status & IUCN status & IUCN Red List \url{http://www.iucnredlist.org}  \\
phylomatic\_tree & Get a plant Phylogeny & Phylomatic \cite{webb2005}  \\
plantminer & Search Plantminer & Plantminer \cite{carvalho2010plantminer}   \\
searchbycommonname & Search ITIS by common name & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
searchbyscientificname & Search ITIS by scientific name & Integrated Taxonomic Information System \url{http://www.itis.gov/}  \\
tax\_name & Get taxonomic name for specific rank & Various  \\
tax\_rank & Get rank of a taxonomic name & Various  \\
tnrs & Resolve names using iPlant & iPlant Taxonomic Name Resolution Service \url{http://tnrs.iplantcollaborative.org/}  \\
tp\_acceptednames & Check for accepted names using Tropicos & Tropicos \url{http://www.tropicos.org/}  \\
tpl\_search & Search the Plant List & The Plant List \url{http://www.theplantlist.org}  \\
ubio\_namebank & Search uBio & uBio \url{http://www.ubio.org/index.php?pagename=sample_tools}  \\
\hline
\end{tabular}
\label{tab:a}
\end{table*}

@@

\section*{Data sources and package details}
taxize uses many data sources (Table~\ref{tab:a}), and more can be easily added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomatic) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize does not venture deeply into these other domains.

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key. This is necessary in some cases so that API providers can 1) better prevent databases crashing from too many requests, 2) collect analytics on requests to their API to provide better performance, etc., and 3) provide user level modification of rules for interacting with the API. The services that require an API key in taxize are: Encyclopedia of Life (EOL) \url{http://eol.org/}, the Universal Biological Indexer and Organizer (uBio) \url{http://www.ubio.org/index.php?pagename=sample_tools}, Tropicos \url{http://www.tropicos.org/}, and Plantminer \cite{carvalho2010plantminer}. One can easily obtain API keys by visiting the website of each service (see Table~\ref{tab:a} for links to each site). There are two typical ways of using API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio\_namebank(srchName='Ursus americanus', key='your\_alphanumeric\_key')}). Second, you can store your key in the .Rprofile file, which is a common place to store settings. We recommend the second option as it simplifies function calls as taxize detects the stored keys.

One available data source in taxize is The Plant List \url{http://www.theplantlist.org}. The connection in taxize is done via the \emph{taxonstand} package \cite{Cayuela_2012,taxonstand} that solely interacts with the Plant List. We provide a few convenience functions that wrap \emph{taxonstand} into taxize.

@@

<<installtaxizecran, eval=FALSE, size='scriptsize'>>=
install.packages("taxize")
@

<<loadtaxize, size='scriptsize'>>=
library(taxize)
@

Advanced users can also download and install the latest development copy from GitHub \url{https://github.com/ropensci/taxize_}, also permanently available at \url{http://dx.doi.org/10.5281/zenodo.7097}.

@@

This is a common task in biology. We often have a list of species names and we want to know a) if we have the most up to date names, b) if our names are spelled correctly, and c) the scientific name for a common name. One way to resolve names is via the Global Names Resolver (GNR) service provided by the Encyclopedia of Life \url{http://resolver.globalnames.org/}. Here, on can search for two misspelled names:

<<resolvenames, results='hide', cache=TRUE, size='scriptsize'>>=
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"))
temp[ , -c(1,4)]
@

<<resolvenames1, echo=FALSE, cache=TRUE, size='scriptsize'>>=
temp[ , -c(1,4)]
@

The correct spellings are \emph{Helianthus annuus} and \emph{Homo sapiens}. Another approach uses the Taxonomic Name Resolution Service via the Taxosaurus API \url{http://taxosaurus.org/} developed by iPLant and the Phylotastic organization. In this example is a list of species names, some of which are misspelled, and then call the API with the \emph{tnrs} function.

<<tnrs, results='hide', cache=TRUE, size='scriptsize'>>=
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
  	"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")
tnrs(query = mynames)[ , -c(5:7)]
@

<<tnrs1, echo=FALSE, cache=TRUE, size='scriptsize'>>=
tnrs(query = mynames)[ , -c(5:7)]
@

@@

Another common use case is when there are many synonyms for a species. In this example, there are six synonyms of the currently accepted name for a species.

<<synonyms, results='hide', cache=TRUE, message=FALSE, warning=FALSE, comment=NA, size='scriptsize'>>=
library(plyr)
mynames <- c("Helianthus annuus ssp. jaegeri", "Helianthus annuus ssp. lenticularis", "Helianthus annuus ssp. texanus", "Helianthus annuus var. lenticularis", "Helianthus annuus var. macrocarpus", "Helianthus annuus var. texanus")
tsn <- get_tsn(mynames)
ldply(tsn, itis_acceptname)
@

<<synonyms1, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA, size='scriptsize'>>=
library(plyr)
mynames <- c("Helianthus annuus ssp. jaegeri", "Helianthus annuus ssp. lenticularis", "Helianthus annuus ssp. texanus", "Helianthus annuus var. lenticularis", "Helianthus annuus var. macrocarpus", "Helianthus annuus var. texanus")
tsn <- get_tsn(mynames)
ldply(tsn, itis_acceptname)
@

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: Integrated Taxonomic Information System (ITIS) \url{http://www.itis.gov/} and National Center for Biotechnology Information (NCBI) \cite{federhen2012}. First, search for two species, \emph{Abies procera} and \emph{Pinus contorta} within ITIS.

<<itisclass, results='hide', cache=TRUE, size='scriptsize'>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(specieslist, db = 'itis')
@

<<itisclass1, echo=FALSE, cache=TRUE, size='scriptsize'>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(specieslist, db = 'itis')
@

@@

<<taxname, results='hide', cache=TRUE, size='scriptsize'>>=
tax_name(query = "Helianthus annuus", get = "family", db = "itis")
tax_name(query = "Helianthus annuus", get = "family", db = "ncbi")
@

<<taxname1, echo=FALSE, cache=TRUE, size='scriptsize'>>=
tax_name(query = "Helianthus annuus", get = "family", db = "itis")
tax_name(query = "Helianthus annuus", get = "family", db = "ncbi")
@

@@

<<interactive, results='hide', cache=TRUE, size='scriptsize'>>=
get_tsn(searchterm = "Heliastes", searchtype = "sciname")
@

<<interactive1, echo=FALSE, cache=TRUE, size='scriptsize'>>=
get_tsn(searchterm = "Heliastes", searchtype = "sciname")
@

@@

<<interactive_many, results='hide', cache=TRUE, size='scriptsize'>>=
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
  	"shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm = splist, searchtype = "sciname")
@

<<interactive_many1, echo=FALSE, cache=TRUE, size='scriptsize'>>=
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
    "shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm = splist, searchtype = "sciname")
@

@@

<<interactive_na, results='hide', cache=TRUE, size='scriptsize'>>=
get_uid(sciname = c("Chironomus riparius", "aaa vva"))
@

<<interactive_na1, echo=FALSE, cache=TRUE, size='scriptsize'>>=
get_uid(sciname = c("Chironomus riparius", "aaa vva"))
@

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, functional trait ecology \cite{poff2006functional}. Yet, Many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in figure \ref{fig:phylomatic}.

<<phylomaticphylogeny_disp, eval=FALSE, size='scriptsize'>>=
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")
tree <- phylomatic_tree(taxa = taxa)
tree$tip.label <- capwords(tree$tip.label)
plot(tree, cex=1)
@

<<phylomaticphylogeny, eval=TRUE, dev=c ('pdf','jpeg'), dpi=300, out.width='4in', out.height='4in', cache=TRUE, include=FALSE, size='scriptsize'>>=
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")
tree <- phylomatic_tree(taxa = taxa)
tree$tip.label <- capwords(tree$tip.label)
plot(tree, cex=1)
@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=3in]{phylogeny.pdf}
\end{center}
\caption{
{A phylogeny for three species. This phylogeny was produced using the \emph{phylomatic\_tree} function, which queries the Phylomatic database, and prunes a previously created phylogeny of plants.}
}
\label{fig:phylomatic}
\end{figure}
